The Categorical Structure of Profunctor Optics

### 2017-08-31 paper.tex
Small edits

\begin{proposition}
\label{prop-profunctor-optics-are-optics}
\begin{align*}
[\Tamb_\C(\M), \Set]((U-)(A,A),(U-)(S,S)) &\cong \PreOptic_\M((S, S), (A, A))
\end{align*}
\end{proposition}
\begin{proof}
We have the chain of isomorphisms:
\begin{align*}
&[\Tamb_\C(\M), \Set]((U-)(A,A),(U-)(S,S)) \\
\cong \;&[\Tamb_\C(\M), \Set](\Tamb_\C(\M)(\Pastro_\M E_A, -), (U-)(S,S))  && \text{(by Lemma \ref{lemma-rep})}\\
\cong \;&(U\Pastro_\M E_A)(S,S)  && \text{(by Yoneda)} \\
= \;&\int^{M \in \M} \int^{C,D \in \C} \C(S, MC) \times E_A(C,D) \times \C(MD,S) \\
= \;&\int^{M \in \M} \int^{C,D \in \C} \C(S, MC) \times \C(C,A) \times \C(A,D) \times \C(MD,S) \\
\cong \;&\int^{M \in \M} \C(S, MA) \times \C(MA, S)  && \text{(by co-Yoneda twice)} \\
= \;&\PreOptic_\M((S, S), (A, A))
\end{align*}
\todo{This works for $\PreOptic_\M((S, S'), (A, A'))$ just fine, just define $E_{A,A'}$ in the obvious way.}
\end{proof}

\begin{corollary}
A profunctor preoptic is completely determined by its component at $\Pastro_\M E_A$, and furthermore, this component is determined by its value on $\langle \id_A, \id_A, \id_A, \id_A \rangle \in (U \Pastro_\M E_A)(A, A)$.
\end{corollary}
\begin{proof}
This is the content of the first two isomorphisms above.
\end{proof}

@@

\begin{proposition}
Any optic $p : (U-)(A,A) \to (U-)(S,S)$ is a lax monoidal transformation with respect to Tambara module composition in the domain and the cartesian product in the codomain.
\end{proposition}
\begin{proof}
The final few isomorphisms used in Proposition \ref{prop-profunctor-optics-are-optics} are essentially window dressing. Suppose we have an element of 
\begin{align*}
(U\Pastro_\M E_A)(S,S) = \int^{M \in \M} \int^{C,D \in \C} \C(S, MC) \times E_A(C,D) \times \C(MD,S)
\end{align*}
and let $\phi : (U\Pastro_\M E_A)(S,S) \cong \PreOptic_\M((S, S), (A, A))$ be the isomorphism given in the proposition. If we have an element of $(U\Pastro_\M E_A)(S,S)$ with representative $\langle l, e_1, e_2, r \rangle$, the round-trip through the isomorphism $\phi$ yields $\langle (M e_1)l, \id_A, \id_A, r(M e_2)\rangle$. Hence we may always assume we have a representative of the form $\langle l, \id_A, \id_A, r \rangle$. This representative corresponds to an optic via the isomorphism $\phi$ iff both $rl = \id_S$ and $\langle lr\rangle = \langle \id_{NA} \rangle$ in $\int^{M \in \M} \C(M A, M A)$. Note that just as we can ``strictify'' a representative of an optic, if we have an element of $(U\Pastro_\M E_A)(S,S)$ that corresponds to an optic we may always choose a representative so that the latter identity holds on the nose.

Now, by the definition of the Yoneda isomorphism, a natural transformation \[ \Tamb_\C(\M)(\Pastro_\M E_A, -) \to (U-)(S,S) \] will correspond to an optic iff the image of the identity morphism in $(U\Pastro_\M E_A)(S, S)$ corresponds to an optic. The very first isomorphism in Proposition \ref{prop-profunctor-optics-are-optics} is given by precomposition by the isomorphism of Lemma \ref{lemma-rep}, so in all, a profunctor preoptic $p : (U-)(A, A) \to (U-)(S,S)$ is a lawful optic iff $p(\langle \id_A, \id_A, \id_A, \id_A \rangle)$ is an optic in $(U\Pastro_\M E_A)(S, S)$. Our goal is to show that this happens exactly when the natural transformation is monoidal.

Let $\langle l, \id_A, \id_A, r \rangle = p(\langle \id_A, \id_A, \id_A, \id_A \rangle)$ be a representative, where $l : S \to M A$ and $r : M A \to S$. We can calculate the component of $p$ at a Tambara module $P$ explicitly:
\begin{align*}
p_P = (UP)(l,r) \alpha_{A,A,M}
\end{align*}
where $\alpha$ is the module structure for $P$.

@@

>>>>

\begin{proposition}
\label{prop-profunctor-optics-are-optics}
\begin{align*}
[\Tamb_\C(\M), \Set]((U-)(A,A),(U-)(S,S)) &\cong \PreOptic_\M((S, S), (A, A))
\end{align*}
\end{proposition}
\begin{proof}
We have the chain of isomorphisms:
\begin{align*}
&[\Tamb_\C(\M), \Set]((U-)(A,A),(U-)(S,S)) \\
\cong \;&[\Tamb_\C(\M), \Set](\Tamb_\C(\M)(\Pastro_\M E_A, -), (U-)(S,S))  && \text{(by Lemma \ref{lemma-rep})}\\
\cong \;&(U\Pastro_\M E_A)(S,S)  && \text{(by Yoneda)} \\
= \;&\int^{M \in \M} \int^{C,D \in \C} \C(S, MC) \times E_A(C,D) \times \C(MD,S) \\
= \;&\int^{M \in \M} \int^{C,D \in \C} \C(S, MC) \times \C(C,A) \times \C(A,D) \times \C(MD,S) \\
\cong \;&\int^{M \in \M} \C(S, MA) \times \C(MA, S)  && \text{(by co-Yoneda twice)} \\
= \;&\PreOptic_\M((S, S), (A, A))
\end{align*}
\todo{This works for $\PreOptic_\M((S, S'), (A, A'))$ just fine, just define $E_{A,A'}$ in the obvious way.}

\todo{Flip the order of the isomorphisms top to bottom?}
\end{proof}

\begin{corollary}
A profunctor preoptic is determined by its component at $\Pastro_\M E_A$, and furthermore, this component is determined by its value on $\langle \id_A, \id_A, \id_A, \id_A \rangle \in (U \Pastro_\M E_A)(A, A)$.
\end{corollary}
\begin{proof}
This is the content of the first two isomorphisms above.

Explicitly, let $p_{\Pastro_\M E_A}(\langle \id_A, \id_A, \id_A, \id_A \rangle) = \langle l, \id_A, \id_A, r \rangle$, where $l : S \to M A$ and $r : M A \to S$. Then the component of $p$ at any other Tambara module $P$ is given by
\begin{align*}
p_P = (UP)(l,r) \alpha_{A,A,M}
\end{align*}
where $\alpha$ is the module structure for $P$.
\end{proof}

@@

\begin{proposition}
If $p : S \hto A$ is an optic then the associated natural transformation $(U-)(A,A) \to (U-)(S,S)$ is a lax monoidal.
\end{proposition}
\begin{proof}
The final few isomorphisms used in Proposition \ref{prop-profunctor-optics-are-optics} are essentially window dressing. Suppose we have an element $\langle l, e_1, e_2, r \rangle$ of 
\begin{align*}
(U\Pastro_\M E_A)(S,S) = \int^{M \in \M} \int^{C,D \in \C} \C(S, MC) \times E_A(C,D) \times \C(MD,S).
\end{align*}
The round trip through the isomorphism $\Phi : (U\Pastro_\M E_A)(S,S) \cong \PreOptic_\M((S, S), (A, A))$ demonstrates that $\langle l, e_1, e_2, r \rangle = \langle (M e_1)l, \id_A, \id_A, r(M e_2)\rangle$. Hence we may always assume we have a representative of the form $\langle l, \id_A, \id_A, r \rangle$. This representative corresponds to an optic via the isomorphism $\Phi$ if both $rl = \id_S$ and $\langle lr\rangle = \langle \phi A \rangle$ in $\int^{M \in \M} \C(M A, M A)$. Note that just as we can ``strictify'' a representative of an optic, if we have an element of $(U\Pastro_\M E_A)(S,S)$ that corresponds to an optic we may always choose a representative so that the latter identity holds on the nose.

Next we examine the image of the optics through the Yoneda isomorphism \[ (U\Pastro_\M E_A)(S,S) \cong [\Tamb_\C(\M), \Set](\Tamb_\C(\M)(\Pastro_\M E_A, -), (U-)(S,S)) \] 
By the definition of the Yoneda isomorphism, a natural transformation \[ p : \Tamb_\C(\M)(\Pastro_\M E_A, -) \to (U-)(S,S) \] will correspond to an optic iff the image of the identity morphism $\id : \Pastro_\M E_A \to \Pastro_\M E_A$ through the component $p_{\Pastro_\M E_A}$ is an optic in $(U\Pastro_\M E_A)(S, S)$.

Finally the isomorphism \[ [\Tamb_\C(\M), \Set](\Tamb_\C(\M)(\Pastro_\M E_A, -), (U-)(S,S)) \cong \; [\Tamb_\C(\M), \Set]((U-)(A,A),(U-)(S,S)) \] is given by precomposition by the isomorphism of Lemma \ref{lemma-rep}, so in summary, a ``profunctor preoptic'' $p : (U-)(A, A) \to (U-)(S,S)$ is a lawful optic iff $p_{\Pastro_\M E_A}(\langle \id_A, \id_A, \id_A, \id_A \rangle)$ is an optic in $(U\Pastro_\M E_A)(S, S)$.

@@

\section{Lenses in Linear Types}
